---
title: "Introduction to Statistical Modelling with `R`"
author: "Dae-Jin Lee < dlee@bcamath.org >"
date: "BCAM-UPV/EHU Courses 2018-2019"
output: 
  slidy_presentation: 
    fig_caption: yes
    font_adjustment: +3
    highlight: monochrome
    keep_md: yes
    theme: cerulean
---

# Introduction to basic programming in `R`

In this Chapter 

# Operators

`R`'s binary and logical operators will look very familiar to programmers. Note that binary operators work on vectors and matrices as well as scalars. 


## Arithmetic Operators 

  Operator      | Description
--------------- | -------------
      `+`       | addition
      `-`       | subtraction
      `*` 	    | multiplication
      `/` 	    | division
  `^` or `**` 	| exponentiation
    x `%%` y 	  | modulus (x mod y) 5%%2 is 1
    x `%/%` y 	| integer division 5%/%2 is 2 

## Logical Operators

  Operator      | Description
--------------- | -------------
      `<`       | less than
      `>`       | greater than
      `<=` 	    | less or equal to
      `>=` 	    | greater or equal to
      `==`      | exactly equal to
      `!=` 	    | not equal to 
      `!x` 	    | Not x
     `x|y`      | `x` OR `y`
     `x&y` 	    | `x` AND `y` 
  `isTRUE(x)` 	|  test if `x` is `TRUE`
      
      
      
```{r}
# An example
x <- c(1:10)
x[(x>8) | (x<5)]
# yields 1 2 3 4 9 10

# How it works
x <- c(1:10)
x
x > 8
x < 5
x > 8 | x < 5
x[c(T,T,T,T,F,F,F,F,T,T)]
```

## Control Structures

These allow you to control the flow of execution of a script typically inside of a function. Common ones include:

  
  +    if, else
  
  +    for
  
  +    while
  
  +    repeat
  
  +    break
  
  +    next
  
  +    return


### Conditional Executions

**`if`**

```{r,eval=FALSE}
if (condition) {
    # do something
} else {
    # do something else
}
```

e.g.:

```{r}
x <- 1:15
if (sample(x, 1) <= 10) { # ?sample  
    print("x is less than 10")
} else {
    print("x is greater than 10")
}
```

**Vectorization with `ifelse`**

```{r,eval=FALSE}
ifelse(x <= 10, "x less than 10", "x greater than 10")
```

Other valid ways of writing if/else


```{r,eval=FALSE}
if (sample(x, 1) < 10) {
    y <- 5
} else {
    y <- 0
}
```

```{r,eval=FALSE}
y <- if (sample(x, 1) < 10) {
    5
} else {
    0
}
```

**for**

A `for` loop works on an iterable variable and assigns successive values till the end of a sequence.

```{r}
for (i in 1:10) {
    print(i)
}
```

```{r}
x <- c("apples", "oranges", "bananas", "strawberries")

for (i in x) {
    print(x[i])
}

for (i in 1:4) {
    print(x[i])
}

for (i in seq(x)) {
    print(x[i])
}

for (i in 1:4) print(x[i])
```


**nested lopps**

```{r}
m <- matrix(1:10, 2)
for (i in seq(nrow(m))) {
    for (j in seq(ncol(m))) {
        print(m[i, j])
    }
}
```

**while**

```{r}
i <- 1
while (i < 10) {
    print(i)
    i <- i + 1
}
```
Be sure there is a way to exit out of a `while` loop.

**Repeat and break**
```{r,eval=FALSE}
repeat {
    # simulations; generate some value have an expectation if within some range,
    # then exit the loop
    if ((value - expectation) <= threshold) {
        break
    }
}
```

**Next**
```{r}
for (i in 1:20) {
    if (i%%2 == 1) {  # %% is the modulus
        next
    } else {
        print(i)
    }
}
```

#### Comparison Operators

  - equal: `==`
  
```{r}
  "hola" == "hola"
  "hola" == "Hola"
   1 == 2-1
```

  
  
  - not equal: `!=`
  

```{r}
    a <- c(1,2,4,5)
    b <- c(1,2,3,5) 
    a == b
    a != b
```
  
  - greater/less than: `>` `<`

```{r}
set.seed(1)
a <- rnorm(10)
b <- rnorm(10)
a<b
```  

  - greater/less than or equal: `>=` `<=`
  
```{r}
set.seed(2)
a <- rnorm(10)
b <- rnorm(10)
a >= b
```


  - `which`
```{r}
set.seed(3)
which(a>b)

LETTERS
which(LETTERS=="R")
```
 
  - `which.min` or `which.max`
```{r}
set.seed(4)
a <- rnorm(10)
a
which.min(a)
which.max(a)
```

  - `is.na`
```{r}
 a[2] <- NA
is.na(a)
which(is.na(a))
```

### Logical Operators

  - and: `&`

```{r}
z = 1:6
which(2 < z & z > 3)
```

  - or: `|`
  
```{r}
z = 1:6
(z > 2) & (z < 5)
which((z > 2) & (z < 5))
```

  - not: `!`

```{r}
x <- c(TRUE,FALSE,0,6)
y <- c(FALSE,TRUE,FALSE,TRUE)

!x
```

Operators `&` and `|` perform element-wise operation producing the result having the length of the longer operand. But `&&` and `||` examines only the first element of the operands resulting into a single length logical vector. Zero is considered `FALSE` and non-zero numbers are taken as `TRUE`.

 **Example:**
  - `&&` vs `&`

```{r}
x&y
x&&y
```

 - `||` vs `|`
 
 
```{r}
x||y
x|y
```


## `if` statements

`if(cond1=true) { cmd1 } else { cmd2 }`

```{r}
if(1==0) {
    print(1)
} else {
    print(2)
}
```

## `ifelse` statement

`ifelse(test, true_value, false_value)`

```{r}
x <- 1:10 # Creates sample data
ifelse(x<5 | x>8, x, 0)
```


## `while` statement


## Loops

The most commonly used loop structures in `R` are `for`, `while` and `apply` loops. Less common are `repeat` loops. The `break` function is used to break out of loops, and next halts the processing of the current iteration and advances the looping index.


### Writing a simple for loop in `R`

Suppose you want to do several printouts of the following form: The year is [year] where [year] is equal to 2010, 2011, up to 2015. You can do this as follows:


```{r}
print(paste("The year is", 2010))
```



### `for`

For loops are controlled by a looping vector. In every iteration of the loop one value in the looping vector is assigned to a variable that can be used in the statements of the body of the loop. Usually, the number of loop iterations is defined by the number of values stored in the looping vector and they are processed in the same order as they are stored in the looping vector.


Syntax
```
for(variable in sequence) {
    statements
}
```


```{r}
for (j in 1:5)
{
  print(j^2)
}
```

Repeat the loop saving the resuls in a vector `x`. 

```{r}
n = 5
x = NULL  # creates a NULL object
for (j in 1:n)
{
  x[j] = j^2
}
x
```

Let's use a for loop to estimate the average of squaring the result of a roll of a dice.

```{r}
nsides = 6
ntrials = 1000
trials = NULL
for (j in 1:ntrials)
{
  trials[j] = sample(1:nsides,1)  # We get one sample at a time
}
mean(trials^2)
```

**Example:** stop on condition and print error message

```{r,eval=FALSE}
x <- 1:10
z <- NULL
for(i in seq(along=x)) {
    if (x[i]<5) {
        z <- c(z,x[i]-1) 
    } else {
        stop("values need to be <5")
    }
}
## Error: values need to be <5
z
## [1] 0 1 2 3
```



## `while`

Similar to `for` loop, but the iterations are controlled by a conditional statement.

```{r}
z <- 0
while(z < 5) {
    z <- z + 2
    print(z) 
}
```

## `apply` loop family


For Two-Dimensional Data Sets: apply

**Syntax:**

```
apply(X, MARGIN, FUN, ARGs)
```

`X`: `array`, `matrix` or `data.frame`; `MARGIN`: 1 for rows, 2 for columns, `c(1,2)` for both; `FUN`: one or more functions; `ARGs`: possible arguments for function.

```{r, eval=FALSE}
## Example for applying predefined mean function
apply(mtcars[,1:3], 1, mean)

## With custom function
x <- 1:10
test <- function(x) { # Defines some custom function
    if(x < 5) {
        x-1
    } else {
        x / x
    }
} 

apply(as.matrix(x), 1, test) 

## Same as above but with a single line of code
apply(as.matrix(x), 1, function(x) { if (x<5) { x-1 } else { x/x } })
```

**For Ragged Arrays: `tapply`**

Apply a function to each cell of a ragged array, that is to each (non-empty) group of values given by a unique combination of the levels of certain factors.

```{r,eval=TRUE}
## Computes mean values of vector agregates defined by factor
tapply(as.vector(mtcars$mpg), factor(mtcars$cyl), mean)

## The aggregate function provides related utilities
aggregate(mtcars[,c(1,3,4)], list(mtcars$cyl), mean)
```


**For Vectors and Lists: `lapply` and `sapply`**

Both apply a function to vector or list objects. The function `lapply` returns a list, while `sapply` attempts to return the simplest data object, such as `vector` or `matrix` instead of `list`. 

*Syntax*

```
lapply(X,FUN)
sapply(X,FUN)
```

```{r, echo=TRUE}
## Creates a sample list
mylist <- as.list(mtcars[,c(1,4,6)])
mylist
```

Compute sum of each list component and return result as list

```{r}
lapply(mylist, sum)
```

Compute sum of each list component and return result as vector

```{r}
sapply(mylist, sum)
```

## Other Loops

**Repeat Loop**

*Syntax*

`repeat` statements

Loop is repeated until a break is specified. This means there needs to be a second statement to test whether or not to break from the loop.

*Example:*
```
z <- 0
repeat {
    z <- z + 1
    print(z)
    if(z > 100) break()
}
```

## Improving Speed Performance of Loops

Looping over very large data sets can become slow in `R`. However, this limitation can be overcome by eliminating certain operations in loops or avoiding loops over the data intensive dimension in an object altogether. The latter can be achieved by performing mainly vector-to-vector or matrix-to-matrix computations which run often over 100 times faster than the corresponding `for()` or `apply()` loops in `R`. For this purpose, one can make use of the existing speed-optimized R functions (e.g.: `rowSums`, `rowMeans`, `table`, `tabulate`) or one can design custom functions that avoid expensive `R` loops by using vector- or matrix-based approaches. Alternatively, one can write programs that will perform all time consuming computations on the C-level.


1. Speed comparison of `for` loops with an append versus and inject step

```{r,eval=FALSE}
N <- 1e3
myMA <- matrix(rnorm(N), N, 10, dimnames=list(1:N, paste("C", 1:10, sep="")))
results <- NULL
system.time(for(i in seq(along=myMA[,1])) 
            results <- c(results, mean(myMA[i,])))

results <- numeric(length(myMA[,1]))
system.time(for(i in seq(along=myMA[,1])) 
            results[i] <- mean(myMA[i,]))
```
The inject approach is 20-50 times faster than the append version.

2. Speed comparison of `apply` loop versus `rowMeans` for computing the mean for each row in a large matrix:

```{r,eval=FALSE}
system.time(myMAmean <- apply(myMA, 1, mean))
system.time(myMAmean <- rowMeans(myMA))
```

The `rowMeans` approach is over 200 times faster than the `apply` loop.




<!-- http://manuals.bioinformatics.ucr.edu/home/programming-in-r#TOC-Control-Structures -->




```{r,eval=FALSE,echo=FALSE,message=FALSE}
library(knitr)
purl("Ch3.Rmd",output="Ch3.R")
```


