---
title: '**Introduction to Statistical Modelling in `R`**'
author: "Dae-Jin Lee < dlee@bcamath.org >"
date: "BCAM-UPV/EHU Courses 2018-2019"
output:
  slidy_presentation:
    fig_caption: yes
    font_adjustment: 5
    highlight: monochrome
    smart: no
    theme: cerulean
  beamer_presentation:
    colortheme: seahorse
    fonttheme: professionalfonts
  ioslides_presentation:
    highlight: monochrome
---


***********

# Analyzing data in `R`

* We are studying inflammation in patients who have been given a new treatment for arthritis.

* The data sets are stored in *comma-separated values* (CSV) format. Each row holds the observations for just one patient. 

* Each column holds the inflammation measured in a day, so we have a set of values in successive days. 

* Download data [here]("http://idaejin.github.io/courses/R/2018/data/inflammation-01.csv")


## Aims

  + Load data into memory.
  
  + Calculate the average value of inflammation per day across all patients, and
  
  + Plot results.
  
# Loading Data


  + Let's import `inflammation-01.csv` into our `R` environment.
  
```{r,eval=FALSE}
setwd("~/.../data/")
```


  + Read file
```{r, results='hide'}
read.csv(file = "data/inflammation-01.csv", header = FALSE)
```

 + Check the difference with 
 
```{r, eval = FALSE}
read.csv(file = "data/inflammation-01.csv") # by default is set to header = TRUE
```
 
